---
layout:  post
title: "Web scraping and webcomics"
comments:  true
published:  true
author: "Zach Burchill"
date: 2015-12-12 20:00:00
categories: ['web scraping',webcomics,python,threading,R]
output:
  html_document:
    mathjax:  default
    fig_caption:  true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

I like learning statistical modeling. I like webcomics.  I 

It started with the thought: "Can I use statistical modelling to tell me when I should stop hoping a webcomic will keep updating?"  There are few feelings worse than returning to a comic week after week in the vain hope its creator will start releasing new strips again.

And while that model is still in the works, I've gotten my hands on a bunch of cool data in the meantime.

## Web scraping

In order to make a model, you generally need data to train it on.  Most webcomics don't have a downloadable R database of all their updates, so you'll have to get the data yourself.  You can click through potentially thousands of pages, recording the dates manually, or you can use **web scraping** to do it automatically.

If you use Python and want an incredibly easy-to-get-into web scraping tool, check out the Python module, [Beautiful Soup](http://www.crummy.com/software/BeautifulSoup/) (I used `bs4`). In a nutshell, it loads HTML pages and parses the elements into a tree automatically. Extracting, say, all the `img` tags on a page can be as simple as: `soup.find_all('img')`. Beautiful Soup is pretty simple to learn and the [documentation](http://www.crummy.com/software/BeautifulSoup/bs4/doc/) makes it a breeze. 

## Data

## Broodhollow updates: a story of an author's life

"Broodhollow."  The name sounds like a title that a moody fifteen-year-old would come up with, but it is **absolutely** my favorite webcomic. _**Ever.**_ The art is beautiful and the story writing is some of the best I've ever read. One part comedy and two parts creeping horror, if author [Kris Straub](http://studios.chainsawsuit.com/) can finish the comic with a _tenth_ of the talent he's exhibited so far, I'm confident this will go down as one of the great graphic novels of our generation.

```{r, echo=FALSE, fig.width=3}
library(png)
library(grid)
img <- readPNG("/Users/zburchill/Desktop/broodhollow_snippet.png")
grid.raster(img)
```

<p class = "figcaption">A sample of Kris Staub's genius.</p>
 
That is, _if_ he can finish it.  Kris is 

## Notes on my code:


